Bayesian non-parametric inference for multivariate peaks-over-threshold models

### 2023-06-15 pointbypoint.tex

    \bruno{Well, here we have to bite the bullet. Hopefully there some software that allows us
    to fit the empirical angular measure.}

    \item \emph{Experiments}:
        \begin{enumerate}
            \item Computational Evidence missing: The convergence of the Gibbs sampler is not 
            discussed in the paper.

            We thank the reviewers for bringing this omission to our attention.  We have now
            included a discussion of MCMC convergence among the models in section 5.

            As we considered parallel tempering to be necessary in fitting the multivariate-lognormal
            model, as we're using a $d$-dimensional joint proposal step, it is readily apparent that 
            convergence will appear \emph{cleaner} on the log-density plots for the lognormal models 
            relative to the gamma models, as parallel tempering also ameliorates autocorrelation
            issues.  Below are the log-density by iteration plots for the ERA-5 data, on the two 
            unrestricted projected gamma models; left uses the product of gammas centering distribution, 
            right uses the log-normal centering distribution.  The red line indicates the end of the 
            burn-in period, and we only keep every tenth sample after.

            \includegraphics[width=0.45\textwidth]{images/ERA-5-PG-G.pdf}
            \includegraphics[width=0.45\textwidth]{images/ERA-5-PG-LN.pdf}

            We could say that parallel tempering would improve autocorrelation for the gamma model,
            but sampling longer and thinning more would serve the same effect.  And since it also
            reaches convergence faster, we could also simply use less burn-in.

### 2023-06-15 abstract.tex
statistics and computing submission
update

% We consider a constructive definition of the multivariate Pareto that factorizes the 
% random vector into a radial component and an independent angular component; The former
% following a univariate Pareto distribution, and the latter defined on the surface of the 
% positive orthant of the unit hypercube.  In this paper, we  propose a method for
% inferring the distribution of the angular component.  We identify its support 
% as the limit of the positive orthants of the unit $p$--norm spheres,
% and introduce a projected gamma family of
% distributions defined as the projection of a vector of independent gamma random variables
% onto the $p$--norm sphere.  This family serves as a building block for a flexible family
% of distributions obtained as a Dirichlet process mixture of projected gammas.  For 
% model assessment and comparison, we discuss model scoring methods appropriate 
% to distributions on the unit hypercube.  In particular, working with the energy score
% criterion, we develop a kernel metric appropriate to the hypercube that produces a 
% proper scoring rule.  We then present a simulation study to compare 
% different modeling choices using the proposed scoring rules.  Finally we apply our 
% approach to describe the dependence structure of the extreme values of the magnitude 
% of the integrated vapor  transport (IVT), a variable that describes the rate of flow 
% of moisture in the atmosphere  along the coast of California for the years of 1979 
% through 2020.  We find a clear but heterogeneous geographical dependence.

We consider a constructive definition of the multivariate Pareto that factorizes the 
random vector into a radial component and an independent angular component; The former
following a univariate Pareto distribution, and the latter defined on the surface of the 
positive orthant of the unit hypercube.  In this paper, we  propose a method for
inferring the distribution of the angular component.  We identify its support 
as the limit of the positive orthant of the unit $p$--norm spheres,
and introduce a projected gamma family of
distributions defined as the projection of a vector of independent gamma random variables
onto the $p$--norm sphere.  This family serves as a building block for a flexible family
of distributions obtained as a Dirichlet process mixture of projected gammas.  For 
model assessment, we discuss scoring methods appropriate 
to distributions on the unit hypercube.  In particular, working with the energy score
criterion, we develop a kernel metric that produces a 
proper scoring rule, and present a simulation study to compare 
different modeling choices using the proposed scoring rules.  Finally we apply our 
approach to describe the dependence structure of the extreme values of the magnitude 
of the integrated vapor transport (IVT), data describing the flow 
of atmospheric moisture along the coast of California for the years of 1979 
through 2020.  We find clear but heterogeneous geographical dependence.

>>>>

We consider a constructive definition of the multivariate Pareto that factorizes the 
random vector into a radial component and an independent angular component; The former
following a univariate Pareto distribution, and the latter defined on the surface of the 
positive orthant of the infinity norm unit hypercube.  In this paper, we  propose a method 
for inferring the distribution of the angular component.  We identify its support 
as the limit of the positive orthant of the unit $p$--norm spheres, and introduce a 
projected gamma family of distributions defined through the normalization a vector of 
independent gamma random variables to the $p$--norm sphere.  This family serves as a 
building block for a flexible family of distributions obtained as a Dirichlet process 
mixture of projected gammas.  For model assessment, we discuss scoring methods appropriate 
to distributions on the unit hypercube.  In particular, working with the energy score
criterion, we develop a kernel metric that produces a  proper scoring rule, and present 
a simulation study to compare  different modeling choices using the proposed scoring rules.  
Finally we apply our  approach to describe the dependence structure of the extreme values of 
the magnitude  of the integrated vapor transport (IVT), data describing the flow 
of atmospheric moisture along the coast of California for the years of 1979 
through 2020.  We find clear but heterogeneous geographical dependence.

### 2023-06-15 evaluation.tex
update

For a finite $p$, the shortest connecting path between two points in ${\mathbb S}_p^{d-1}$
  is the minimum geodesic; its length satisfying the definition of a distance.  Thus its
  length can be used as a negative definite kernel for the purpose of defining an energy
  score. Unfortunately as $p\to\infty$ the resulting surface ${\mathbb S}_{\infty}^{d-1}$
  is not differentiable, implying that routines to calculate geodesics are not readily
  available.  However, as ${\mathbb S}_{\infty}^{d-1}$ is a portion of a $d$-cube, we
  can borrow a result from geometry \citep{pappas1989} stating that the length of the
  shortest path between two points on a geometric figure corresponds to the length of a
  straight line drawn between the points on an appropriate unfolding, rotation, or 
  \emph{net} of the figure from a $d$-dimensional to a $d-1$-dimensional space.  The optimal 
  net will have the shortest straight line between the points, as long as that line is fully 
  contained within such net. As ${\mathbb S}_{\infty}^{d-1}$ has $d$ faces---each face 
  pairwise adjacent, there are $d!$ possible nets.  However, we are only interested in nets 
  that begin and end on the source and destination faces respectively, reducing the number 
  of nets under consideration to $\sum_{k = 0}^{d-2}\binom{d-2}{k}$.  This is still 
  computationally burdensome for a large number of dimensions.  
  \makenote{However, we can efficiently establish an upper bound on the geodesic 
  length.  We use this upper bound on geodesic distance as the kernel function for
  the energy score.}

To calculate the energy score we define the kernel 
    \[  
    g(\bm{a},\bm{b}) = \min_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
    \]
    where $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and 
    $\bm{b} \in {\mathbb   C}_{\jmath}^{d-1}$, for $\ell,\jmath\in \{1,\ldots, d\}$.
Evaluating $g$ as described requires the solution of a $(d-2)$-dimensional 
  optimization problem.  The following proposition provides a straightforward approach.
\begin{prop}\label{prop:rot}
    Let $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-1}$, 
        for $\ell, \jmath \in \{1, \ldots , d\}$.  For $\ell\neq \jmath$, the 
  transformation $P_{\jmath\ell}(\cdot)$ required to rotate the $\jmath$th face along the 
  $\ell$th axis produces the vector $\bm{b}^\prime$, where
  \begin{equation}
    \label{eqn:rotation}
     \bm{b}^{\prime}_i = P_{\jmath\ell}(\bm{b}) = 
    \begin{cases}
        b_{i} &\text{for }i\neq \jmath,\ell\\
        1 &\text{for }i = \ell\\
        2 - b_{\ell} &\text{for }i = \jmath
    \end{cases}.
  \end{equation}
  Then $g(\bm{a},\bm{b}) = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}$.
\end{prop}
{\em Proof:}
Notice that for 
  $\bm{c} \in {\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}$,
     $\pnorm{\bm{b} - \bm{c}}{2} =  \pnorm{\bm{b}^{\prime} - \bm{c}}{2}$ and let
\[\bm{e} = 
    \argmin_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
\]
We then have that
  \begin{equation*}
    \begin{aligned}
    g(\bm{a}, \bm{b}) &= \pnorm{\bm{a} - \bm{e}}{2} + \pnorm{\bm{b} - \bm{e}}{2}\\
    &= \pnorm{\bm{a} - \bm{e}}{2} + \pnorm{\bm{b}^{\prime} 
        - \bm{e}}{2} = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}.
    \end{aligned}
  \end{equation*}
  The last equality is due to the fact that $\bm{a}$ and $\bm{b}'$ belong to the 
  same hyperplane. $\hfill\Box$

Using the rotation in Proposition \ref{prop:rot} we obtain the following result.
    \begin{prop}\label{prop:g}
    g is a negative definite kernel.
    \end{prop}
    {\em Proof:} For a given $n$ consider an arbitrary set of points
    $\bm{a}_1, \ldots , \bm{a}_n \in {\mathbb S}_\infty^{d-1}$, and 
    $\alpha_1, \ldots , \alpha_n \in {\mathbb R}$, such that $\sum_{i=1}^n
    \alpha _i = 0$. Then
    \[
    \sum_{i,\jmath} \alpha_i \alpha_\jmath g(\bm{a}_i, \bm{a}_\jmath) 
        = \sum_{i,\jmath} \alpha_i \alpha_\jmath 
    \|\bm{a}_i - \bm{a}_\jmath^\prime\|_2 \leq 0,
    \]
    where $\bm{a}_\jmath^\prime$ is defined as in Proposition \ref{prop:rot}. $\hfill\Box$

>>>>

For a finite $p$, the shortest connecting path between two points in ${\mathbb S}_p^{d-1}$
  is the minimum geodesic; its length satisfying the definition of a distance.  Thus its
  length can be used as a negative definite kernel for the purpose of defining an energy
  score. Unfortunately as $p\to\infty$ the resulting surface ${\mathbb S}_{\infty}^{d-1}$
  is not differentiable, implying that routines to calculate geodesics are not readily
  available.  However, as ${\mathbb S}_{\infty}^{d-1}$ is a portion of a $d$-cube, we
  can borrow a result from geometry \citep{pappas1989} stating that the length of the
  shortest path between two points on a geometric figure corresponds to the length of a
  straight line drawn between the points on an appropriate unfolding, rotation, or 
  \emph{net} of the figure from a $d$-dimensional to a $d-1$-dimensional space.  The optimal 
  net will have the shortest straight line between the points, as long as that line is fully 
  contained within such net. As ${\mathbb S}_{\infty}^{d-1}$ has $d$ faces---each face 
  pairwise adjacent, there are $d!$ possible nets.  However, we are only interested in nets 
  that begin and end on the source and destination faces respectively, reducing the number 
  of nets under consideration to $\sum_{k = 0}^{d-2}\binom{d-2}{k}$.  This is still 
  computationally burdensome for a large number of dimensions.  
  However, we can efficiently establish an upper bound on the geodesic 
  length.  We use this upper bound on geodesic distance as the kernel function for
  the energy score.

To calculate the energy score we define the kernel 
    \[  
    g(\bm{a},\bm{b}) = \min_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
    \]
    where $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and 
    $\bm{b} \in {\mathbb   C}_{\jmath}^{d-1}$, for $\ell,\jmath\in \{1,\ldots, d\}$.
Evaluating $g$ as described requires the solution of a $(d-2)$-dimensional 
  optimization problem.  The following proposition provides a straightforward approach.
\begin{prop}\label{prop:rot}
    Let $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-1}$, 
        for $\ell, \jmath \in \{1, \ldots , d\}$.  For $\ell\neq \jmath$, the 
  transformation $P_{\jmath\ell}(\cdot)$ required to rotate the $\jmath$th face along the 
  $\ell$th axis produces the vector $\bm{b}^\prime$, where
  \begin{equation}
    \label{eqn:rotation}
     \bm{b}^{\prime}_i = P_{\jmath\ell}(\bm{b}) = 
    \begin{cases}
        b_{i} &\text{for }i\neq \jmath,\ell\\
        1 &\text{for }i = \ell\\
        2 - b_{\ell} &\text{for }i = \jmath
    \end{cases}.
  \end{equation}
  Then $g(\bm{a},\bm{b}) = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}$.
\end{prop}
{\em Proof:}
Notice that for 
  $\bm{c} \in {\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}$,
     $\pnorm{\bm{b} - \bm{c}}{2} =  \pnorm{\bm{b}^{\prime} - \bm{c}}{2}$. %and let
%\[\bm{e} = 
%    \argmin_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
%        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
%\]
We then have that
  \begin{equation*}
    \begin{aligned}
    g(\bm{a},\bm{b}) = &\min_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\} \\
     &\min_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}^{\prime}-\bm{c}}{2} \right\}\\
    &= \pnorm{\bm{a} - \bm{b}^{\prime}}{2}.
    \end{aligned}
  \end{equation*}
  The last equality is due to the fact that $\bm{a}$ and $\bm{b}'$ belong to the 
  same hyperplane. $\hfill\Box$

Using the rotation in Proposition \ref{prop:rot} we obtain the following result.
    \begin{prop}\label{prop:g}
    g is a negative definite kernel.
    \end{prop}
    {\em Proof:} For a given $n$ consider an arbitrary set of points
    $\bm{a}_1, \ldots , \bm{a}_n \in {\mathbb S}_\infty^{d-1}$, and 
    $\alpha_1, \ldots , \alpha_n \in {\mathbb R}$, such that $\sum_{i=1}^n
    \alpha _i = 0$. Then
    \[
    \sum_{i,\jmath} \alpha_i \alpha_\jmath g(\bm{a}_i, \bm{a}_\jmath) 
        = \sum_{i,\jmath} \alpha_i \alpha_\jmath 
    \|\bm{a}_i - \bm{a}_\jmath^\prime\|_2 \leq 0,
    \]
    where $\bm{a}_\jmath^\prime$ is defined as in Proposition \ref{prop:rot}. 
    The last equality holds as $\|\bm{x} - \bm{x}^\prime\|_2, \bm{x},
    \bm{x}^{\prime} \in {\mathbb R}^d$ is negative definite
    \citep{gneiting2007}$\hfill\Box$

### 2023-06-15 introduction.tex
update

Inferential methods for the extreme values of univariate observations
    are well established and software is widely available 
    \cite[see, for example,][]{coles2001}. For variables in one dimension the 
    application of EVT methods considers the asymptotic distribution of either the 
    maxima calculated for regular blocks of data, or the values that exceed a 
    certain threshold. The former leads to a Generalized Extreme Value (GEV) 
    distribution, that depends on three parameters. The latter leads to a Generalized 
    Pareto (GP) distribution, that depends on a shape and a scale parameter. 
    Likelihood-based approaches to inference can be readily implemented in both cases.
    In the multivariate case the GEV theory is well developed 
    \citep[see, for example][]{dehaan2006}, but the inferential problem is more 
    complicated, as there is no parametric representation of the GEV. This problem
    is inherited by the peaks over threshold (PoT) approach and compounded by the 
    fact that there is no unique definition of an exceedance of a multivariate 
    threshold, as there is an obvious dependence on the norm that is used to measure 
    the size of a vector.

During the last decade or so, much work has been done in the exploration of the 
    definition and properties of an appropriate generalization of the univariate 
    GP distribution to a multivariate setting.  To mention some of the papers on 
    the topic, the work of \citep{rootzen2006} defines the generalized Pareto 
    distribution, with further analysis on these classes of distributions 
    presented in \cite{falk2008} and \cite{michel2008}.  A recent review of the 
    state of the art in multivariate peaks over threshold modelling using 
    generalized Pareto is provided in \cite{rootzen2018} while \cite{RoSeWa2018a} 
    provides insight on the theoretical properties of possible parametrizations. 
    These are use in \cite{KiRoSeWa2019} for likelihood-based models for PoT 
    estimation. A frequently used method for describing the dependence in 
    multivariate distributions is to use a copula. \cite{renard2007}, and 
    \cite{falk2019} provide successful  examples of this approach in an EVT 
    framework. \cite{ferreira2014} presents a constructive definition of the 
    Pareto process, that generalizes the GP to an infinite dimensional setting. It 
    consists of decomposing the process into independent radial and angular 
    components. Such an approach can be used in the finite dimensional case, where 
    the angular component contains the information pertaining to the dependence 
    structure of the random vector. Based on this definition, we present a novel 
    approach for modelling  the angular component with families of distributions 
    that provide flexibility and can be applied in a moderately large dimensional 
    setting.  Our focus on the angular measure is similar to that in \cite{boldi2007},
    \cite{SaNa2014} and \cite{HaCaCh2017}, that consider Bayesian non-parametric 
    approaches. Yet, our approach differs in that it is established in the 
    peaks-over-threshold regime, and avoids the problem of dealing with the so 
    called moment constraint by considering a constructive definition of the 
    multivariate GP, based on the infinity norm.  Further, the

@@

\section{A multivariate PoT model\label{sec:multivariatepot}}
To develop a multivariate PoT model for extreme values consider a 
    $d$-dimensional random vector $\bm{W} = (W_1, \ldots ,W_d)$ with
    cumulative distribution $F$. Following \cite{RoSeWa2018a}, assume 
    that there exists sequences of vectors $\bm{a}_n$ and $\bm{b}_n$,
    and a $d$-variate distribution $G$ such that 
    $\lim_{n\rightarrow\infty} F^n(\bm{a}_n \bm{w} + \bm{b}_n) = 
    G(\bm{w})$. $G$ is a $d$-variate generalized extreme value 
    distribution. It follows that
    \begin{equation*}
        \begin{aligned}
        \lim\limits_{n\rightarrow\infty} &\prob{\bm{a}^{-1}_n (\bm{W} - \bm{b}_n) 
      \leq \bm{w} \mid \bm{W} \not\leq \bm{b}_n)} \\ 
        &\hspace{0.5cm}= \frac{\log G(\bm{w} \wedge \bm{0}) 
        - \log G(\bm{w}) }{\log G(\bm{0})} = H(\bm{w}).
        \end{aligned}
    \end{equation*}
    $H$ a multivariate Pareto distribution, and corresponds to the joint
    distribution conditional on exceeding a multivariate threshold. In addition to
    a copula function, $H$ depends on two $d$-dimensional vectors of 
    parameters that we denote as $\bm{\xi}$ for the shapes and $\bm{\sigma}$ 
    for the scales. \cite{RoSeWa2018a} provide a number of stochastic 
    representations for $H$.  In particular, denote as $Z$ a
    random variable with distribution $H$ where $\bm{\xi}= 1$ and 
    $\bm{\sigma} = 0$.  Then, Remark 1 justifies the representation in 
    \cite{ferreira2014}, giving $\bm{Z} = \bm{V}R$
    where $R$ and $\bm{V}$ are independent. $R = \|\bm{Z}\|_\infty$ is 
    distributed as a standard Pareto random variable, and $\bm{V} = 
    \bm{Z}/\|\bm{Z}\|_\infty$ is a random vector in  
    $\mathbb{S}_{\infty}^{d-1}$, the positive orthant of the 
    unit sphere under $\mathcal{L}_{\infty}$ norm, with distribution $\Phi$. 
    This representation is central to the methods proposed in this paper.
    $R$ and $\bm{V}$ are referred to, respectively, as the {\em radial} 
    and {\em angular} components of $H$. The angular measure controls 
    the dependence  structure of $\bm{Z}$  in the tails. In view of 
    this, to obtain a  PoT model we seek a flexible model for the 
    distribution of $\bm{V} \in {\mathbb S}_{\infty}^{d-1}$.

>>>>

Inferential methods for the extreme values of univariate observations
    are well established and software is widely available 
    \cite[see, for example,][]{coles2001}. For variables in one dimension the 
    application of EVT methods considers the asymptotic distribution of either the 
    maxima calculated for regular blocks of data, or the values that exceed a 
    certain threshold. The former leads to a Generalized Extreme Value (GEV) 
    distribution, that depends on three parameters. The latter leads to a Generalized 
    Pareto (GP) distribution, that depends on a shape and a scale parameter. 
    Likelihood-based approaches to inference can be readily implemented in both cases.
    In the multivariate case the GEV theory is well developed 
    \citep[see, for example][]{dehaan2006}, but the inferential problem is  
    complicated by the fact that there is no parametric representation of the GEV. 
    This problem is inherited by the peaks over threshold (PoT) approach and 
    compounded by the  fact that there is no unique definition of an exceedance of 
    a multivariate  threshold, as there is an obvious dependence on the norm that 
    is used to measure  the size of a vector.

During the last decade or so, much work has been done in the exploration of the 
    definition and properties of an appropriate generalization of the univariate 
    GP distribution to a multivariate setting.  To mention some of the papers on 
    the topic, the work of \citep{rootzen2006} defines the generalized Pareto 
    distribution, with further analysis on these classes of distributions 
    presented in \cite{falk2008} and \cite{michel2008}.  A recent review of the 
    state of the art in multivariate peaks over threshold modelling using 
    generalized Pareto is provided in \cite{rootzen2018} while \cite{RoSeWa2018a} 
    provides insight on the theoretical properties of possible parametrizations. 
    These are use in \cite{KiRoSeWa2019} for likelihood-based models for PoT 
    estimation. A frequently used method for describing the dependence in 
    multivariate distributions is to use a copula. \cite{renard2007}, and 
    \cite{falk2019} provide successful  examples of this approach in an EVT 
    framework. \cite{ferreira2014} presents a constructive definition of the 
    Pareto process, that generalizes the GP to an infinite dimensional setting. It 
    consists of decomposing the process into independent radial and angular 
    components. Such an approach can be used in the finite dimensional case, where 
    the angular component contains the information pertaining to the dependence 
    structure of the random vector. Based on this definition, we present a novel 
    approach for modelling  the angular component with families of distributions 
    that provide flexibility and can be applied in a moderately large dimensional 
    setting.  Our focus on the angular measure is similar to that in \cite{boldi2007},
    \cite{SaNa2014} and \cite{HaCaCh2017}, that consider Bayesian non-parametric 
    approaches. Yet, our approach differs in that it is established in the 
    peaks-over-threshold regime, and avoids the problem of dealing with the so 
    called moment constraint by considering a constructive definition of the 
    multivariate GP, based on the infinity norm.

@@

\section{A multivariate PoT model\label{sec:multivariatepot}}
To develop a multivariate PoT model for extreme values consider a 
    $d$-dimensional random vector $\bm{W} = (W_1, \ldots ,W_d)$ with
    cumulative distribution $F$. Following \cite{RoSeWa2018a}, assume 
    that there exists sequences of vectors $\bm{a}_n$ and $\bm{b}_n$,
    and a $d$-variate distribution $G$ such that 
    $\lim_{n\rightarrow\infty} F^n(\bm{a}_n \bm{w} + \bm{b}_n) = 
    G(\bm{w})$. $G$ is a $d$-variate generalized extreme value 
    distribution. It follows that
    \begin{equation*}
        \begin{aligned}
        \lim\limits_{n\rightarrow\infty} &\prob{\bm{a}^{-1}_n (\bm{W} - \bm{b}_n) 
      \leq \bm{w} \mid \bm{W} \not\leq \bm{b}_n)} \\ 
        &\hspace{0.5cm}= \frac{\log G(\bm{w} \wedge \bm{0}) 
        - \log G(\bm{w}) }{\log G(\bm{0})} = H(\bm{w}).
        \end{aligned}
    \end{equation*}
    $H$ is a multivariate Pareto distribution, and corresponds to a joint
    distribution conditional on exceeding a multivariate threshold. In addition to
    a copula function, $H$ depends on two $d$-dimensional vectors of 
    parameters that we denote as $\bm{\xi}$ for the shapes and $\bm{\sigma}$ 
    for the scales. \cite{RoSeWa2018a} provide a number of stochastic 
    representations for $H$.  In particular, denote as $Z$ a
    random variable with distribution $H$ where $\bm{\xi}= 1$ and 
    $\bm{\sigma} = 0$.  Then, Remark 1 justifies the representation in 
    \cite{ferreira2014}, giving $\bm{Z} = \bm{V}R$
    where $R$ and $\bm{V}$ are independent. $R = \|\bm{Z}\|_\infty$ is 
    distributed as a standard Pareto random variable, and $\bm{V} = 
    \bm{Z}/\|\bm{Z}\|_\infty$ is a random vector in  
    $\mathbb{S}_{\infty}^{d-1}$, the positive orthant of the 
    unit sphere under $\mathcal{L}_{\infty}$ norm, with distribution $\Phi$. 
    This representation is central to the methods proposed in this paper.
    $R$ and $\bm{V}$ are referred to, respectively, as the {\em radial} 
    and {\em angular} components of $H$. The angular measure controls 
    the dependence  structure of $\bm{Z}$  in the tails. In view of 
    this, to obtain a  PoT model we seek a flexible model for the 
    distribution of $\bm{V} \in {\mathbb S}_{\infty}^{d-1}$.

### 2023-06-15 methodology_projgamma.tex
update

Unfortunately, in the limit when $p\rightarrow \infty$, the projection 
  transformation is not differentiable. Thus, a closed form expression 
  like Equation (\ref{eqn:projgamma}) for the projected gamma density on ${\mathbb S}_\infty^{d-1}$ is not 
  available. Instead, we observe that for a sufficiently large $p$, $\mathbb{S}_p^{d-1}$ will approach
  $\mathbb{S}_{\infty}^{d-1}$.  With that in mind, our strategy consists of describing the angular
  distribution $\Phi$ using a sample based approach with the following steps: (i) Apply the 
  transformation in Equation (\ref{eqn:standardization}) to the original data; (ii) Obtain the 
  subsample of the standardized observations that satisfy $R>1$; (iii) Take a finite $p$ and 
  project the observations onto ${\mathbb S}_p^{d-1}$; 
  (iv) Fit the model in Equation (\ref{eqn:PGmix}) to the resulting 
  data and obtain samples from the fitted model; 
  (v) Project the resulting samples onto ${\mathbb S}_\infty^{d-1}$.
  For step (iv) we use a Bayesian approach that is implemented using a purposely 
  developed Markov chain Monte Carlo sampler described in the next section.

@@

To effectively employ the sample space for a joint Metropolis step, as well as
    speeding convergence, we employ a parallel tempering algorithm\citep{earl2005} 
    for the log-normal models. This technique runs parallel MCMC
    chains at ascending temperatures--that is, for chain $s$, the posterior density is 
    exponentiated to the reciprocal of temperature $t_s$.  For the \emph{cold} chain, 
    $t_1 := 1$.  Let $E_s$ be the log-posterior density under the current parameter 
    state for chain $s$.  Then states between chains $r,s$ are exchanged via a Metropolis 
    step with probability 
    \[
    \alpha_{rs} = \min
    \left[1, \exp\left\lbrace(t_{r}^{-1} - t_{s}^{-1})(E_r - E_s)\right\rbrace\right].
    \]
    Higher temperatures serve to \emph{flatten} the posterior distribution, meaning hotter
    chains have a higher probability of making a given transition, or will make larger 
    transitions.  As such, they will more quickly explore the parameter space, and share
    information gained through state exchange.

>>>>

Unfortunately, in the limit when $p\rightarrow \infty$, the normalizing 
  transformation is not differentiable. Thus, a closed form expression 
  like Equation (\ref{eqn:projgamma}) for the projected gamma density on ${\mathbb S}_\infty^{d-1}$ is not 
  available. Instead, we observe that for a sufficiently large $p$, $\mathbb{S}_p^{d-1}$ will approach
  $\mathbb{S}_{\infty}^{d-1}$.  With that in mind, our strategy consists of describing the angular
  distribution $\Phi$ using a sample based approach with the following steps: (i) Apply the 
  transformation in Equation (\ref{eqn:standardization}) to the original data; (ii) Obtain the 
  subsample of the standardized observations that satisfy $R>1$; (iii) Take a finite $p$ and 
  project the observations onto ${\mathbb S}_p^{d-1}$; 
  (iv) Fit the model in Equation (\ref{eqn:PGmix}) to the resulting 
  data and obtain samples from the fitted model; 
  (v) Project the resulting samples onto ${\mathbb S}_\infty^{d-1}$.
  For step (iv) we use a Bayesian approach that is implemented using a purposely 
  developed Markov chain Monte Carlo sampler described in the next section.

@@

To effectively explore the sample space with a joint Metropolis step, as well as
    to speed convergence, we implement a parallel tempering algorithm\citep{earl2005} 
    for the log-normal models. This technique runs parallel MCMC
    chains at ascending temperatures. That is, for chain $s$, the posterior density is 
    exponentiated to the reciprocal of temperature $t_s$.  For the \emph{cold} chain, 
    $t_1 := 1$.  Let $E_s$ be the log-posterior density under the current parameter 
    state for chain $s$.  Then states between chains $r,s$ are exchanged via a Metropolis 
    step with probability 
    \[
    \alpha_{rs} = \min
    \left[1, \exp\left\lbrace(t_{r}^{-1} - t_{s}^{-1})(E_r - E_s)\right\rbrace\right].
    \]
    Higher temperatures serve to \emph{flatten} the posterior distribution, meaning hotter
    chains have a higher probability of making a given transition, or will make larger 
    transitions.  As such, they will more quickly explore the parameter space, and share
    information gained through state exchange.

### 2023-06-15 results.tex
update

\section{Data illustrations\label{sec:results}} 
We apply the aforementioned models to generated angular data, as well as a data analysis
    of atmospheric data.
    Assessing convergence of an MCMC chain for a large-dimensional model is difficult.
    For our part, we assess convergence by stabilization of the log-posterior 
    density. For all of our datasets, none of the models required more than 40 thousand 
    iterations to achieve stability, so we use that as a burn-in, and thereafter sample 
    an additional 10 thousand iterations. We then thin the chain by 10 to collect 1000 
    total samples with which to generate posterior predictive densities.
    Implementation details play a role here: posterior samples from the gamma-centered 
    models are moderately autocorrelated as compared to the log-normal centered models.
    This is a benefit of the parallel tempering algorithm employed in fitting the 
    log-normal models, but parallel tempering necessitates an additional computational cost,
    growing linearly with the number of temperatures.  We used a temperature ladder
    of $t_s = 1.3^{s}$, for $s \in \lbrace0,1,\ldots,5$.  This was verified to produce
    acceptable swap probabilities both for the simulated data, and real data.

Our hyperprior parameters are set as follows: for the gamma-centered models (PG--G, PRG--G), 
    the shape parameter for the centering 
    distribution~$\xi_{\ell}\sim \mathcal{G}\left(1,1\right)$, and
    rate parameter~$\tau_{\ell}\sim\mathcal{G}\left(2,2\right)$.
    For the log-normal centered models (PG--LN, PRG--LN), the centering distribution's
    log-mean $\bm{\mu}\sim\mathcal{N}_d\left(0,\bm{I}_d\right)$, and covariance 
    matrix~$\Sigma\sim\mathcal{IW}\left(d + 10, (d+10)\bm{I}_{d}\right)$.  These values are 
    intended such that draws from the prior for $\Sigma$ will weakly tend towards the 
    identity matrix.
    For models learning rate parameters $\beta_{j\ell}$ (PG--G, PG--LN), the centering
    distribution's shape parameter $\zeta\sim\mathcal{G}\left(1,1\right)$ and rate parameter
    $\sigma\sim\mathcal{G}\left(2,2\right)$.  The choice of $\mathcal{G}(2,2)$ for rate
    parameters was meant to draw estimates for the value away from 0 for numerical stability.

>>>>

\section{Data illustrations\label{sec:results}} 
We apply the aforementioned models to simulated angular data. We then consider the analysis
    of atmospheric data. To tackle the difficult problem of assessing the convergence 
    an MCMC chain for a large-dimensional model, we monitor the log-posterior 
    density. In all the examples considered MCMC samples produced stable tarces of the
    log-posterior in less than 40 thousand  iterations. Thus, so we use that as a burn-in, 
    and thereafter sample 10 thousand additional iterations. We then thin the chain by retaining one
    every ten samples, to obtain 1000 total samples. These are used to generate samples from the 
    posterior predictive densities. We used two different strategies implemment the MCMC samplers.
    For the models whose DP prior is centered around a log-normal distribution we used parallel 
    tempering. For the models gamma-centered models, we used a single chain. The temperature ladder was set as $t_s = 1.3^{s}$, 
    for $s \in \lbrace0,1,\ldots,5$. This was set empirically in order to produce
    acceptable swap probabilities both for the simulated data, and real data. Our attempts to use
    a single chain produced sample with very slow mixing. On the other hand, from our examples 
    we observe that posterior samples from the gamma-centered models, that do not use parallel 
    tempering, show slightly stronger autocorrelation than those of the log-normal centered 
    models. Parallel tempering produces chains with good mixing properties, but has a computational 
    cost that  grows linearly with the number of temperatures. In view of this, we use it for the

Our hyperprior parameters are set as follows: for the gamma-centered models (PG--G, PRG--G), 
    the shape parameter for the centering 
    distribution~$\xi_{\ell}\sim \mathcal{G}\left(1,1\right)$, and
    rate parameter~$\tau_{\ell}\sim\mathcal{G}\left(2,2\right)$.
    For the log-normal centered models (PG--LN, PRG--LN), the centering distribution's
    log-mean $\bm{\mu}\sim\mathcal{N}_d\left(0,\bm{I}_d\right)$, and covariance 
    matrix~$\Sigma\sim\mathcal{IW}\left(d + 10, (d+10)\bm{I}_{d}\right)$.  These values are 
    intended such that draws from the prior for $\Sigma$ will weakly tend towards the 
    identity matrix.
    For models learning rate parameters $\beta_{j\ell}$ (PG--G, PG--LN), the centering
    distribution's shape parameter $\zeta\sim\mathcal{G}\left(1,1\right)$ and rate parameter
    $\sigma\sim\mathcal{G}\left(2,2\right)$.  The choice of the $\mathcal{G}(2,2)$ for rate
    parameters places no mass near 0, in order to draw estimates for the value away from 
    0 for numerical stability.
